---
title: "Loan word maps"
author:
  - Henning Schreiber:
      correspondence: "yes"
      email: janedoe@fosg.org
      orcid: 0000-0003-1689-0557
      institute:
        - uhh
  - Nikolas Gestrich:
      institute: fi
      orcid: 0000-0003-1689-0558
 
institute:
  - uhh:
      name: Universität Hamburg
      address: 23 Science Street, Eureka, Mississippi, USA
  - fi: Frobenius Institute
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

# Creating language maps

Read location data from glottolog and Sahelia databases

```{r}
#| label: get-data
#| eval: false
# Read data
langloc <- readr::read_csv(here::here('analysis/data/raw_data/language_location.csv')) |>
  
# drop languages with unknown location
  tidyr::drop_na(longitude) |> 

  # turn latitude and longitude columns into geometry
  sf::st_as_sf(coords = c("longitude","latitude"))

# set crs
sf::st_crs(langloc) <- 4326
```

```{r}
# read classified loan word data

# Define the file path
file_path <- here::here('analysis/data/raw_data/cleaned exports.xlsx')

# Get the names of all the sheets in the Excel file
sheet_names <- readxl::excel_sheets(file_path)

# Read each sheet and store it as a separate data frame
for (sheet in sheet_names) {
  assign(sheet, readxl::read_excel(file_path, sheet = sheet, col_names = TRUE)) 
}

```

add the location data 

```{r}
# add location to the gold sheet

# merge by glottolog code 
gold <- gold |> 
  #join by glottolog code
  dplyr::left_join(langloc, by = c("lang" = "GlottCode"))

# split off where glottolog code matching was successful
gold_a <-  gold |> dplyr::filter(!is.na(LanName))


gold_a <- gold |> filter(!is.na(LanName))

#where the glottolog code is not available, merge by Sahelia name
gold <- gold |> 
  dplyr::mutate(id2 = ifelse(is.na(LanName), id2, NA))

# Perform the first join on the primary id1
merged_data <- data1 %>%
  left_join(data2, by = c("id1" = "id"))

# For the rows where the primary ID did not match, perform a second join on id2
final_data <- merged_data %>%
  mutate(id2 = ifelse(is.na(x), id2, NA)) %>%   # Set id2 as NA if the primary join succeeded
  left_join(data2, by = c("id2" = "id")) %>%
  select(-id2)  # Drop the temporary id2 column
```


```{r}
library(dplyr)
library(stringr)
library(sf)
library(readr)
library(leaflet)
library(purrr)
library(glue)
library(htmltools)

# 3. Load classified language data and plot

# Load data
gold <- read_csv("gold.csv") %>%

  # Modify the "lang" column to keep only the first expression and remove the number behind it
  mutate(lang = str_trim(str_extract(lang, "^[^ ]+"))) %>%

  # Join the langloc dataset with the gold dataset
  inner_join(langloc, by = c("lang" = "lang_code")) %>%

  # Exclude rows where the "form" variable contains NA values
  filter(!is.na(form))

# 4. Plot on an interactive leaflet map
# Define a color palette for the "form" column
color_pal <- colorFactor(palette = "Set3", domain = gold$form)

# Add a new column with the corresponding colors
gold <- gold %>%
  mutate(color = color_pal(form))

# Create the leaflet map
leaflet(gold) %>%
  addProviderTiles(providers$Esri.WorldPhysical) %>%
  addLabelOnlyMarkers(
    lng = ~st_coordinates(geometry)[,1],
    lat = ~st_coordinates(geometry)[,2],
    label = purrr::map(
      glue::glue("<span style='color:{gold$color}; font-weight: bold; font-size: 12px;'>{as.character(gold$form)}</span>"),
      htmltools::HTML
    ),
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = 'top',
      textOnly = TRUE
    )
  )


## NEXT: run this for a list of files and output a html file leaflet map for each.

```



Merge using two columns

```{r}
# Perform the first join on the primary id1
merged_data <- data1 %>%
  left_join(data2, by = c("id1" = "id"))

# For the rows where the primary ID did not match, perform a second join on id2
final_data <- merged_data %>%
  mutate(id2 = ifelse(is.na(x), id2, NA)) %>%   # Set id2 as NA if the primary join succeeded
  left_join(data2, by = c("id2" = "id")) %>%
  select(-id2)  # Drop the temporary id2 column
```




```{r}
#| label: fig-demo-plot
#| fig-cap: "A plot of random numbers"
plot(rnorm(10))
```

@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`

```{r}
#| label: demo-inline-code
#| echo: false
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
